---
title: "Assignment 4: Methods of Moment Estimation"
author: "Samuel Johnson"
date: " Due: Febuary, 2026"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: no
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: yes
    theme: lumen
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 3
    fig_height: 3
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
editor_options: 
  chunk_output_type: inline
---

```{css, echo = FALSE}
#TOC::before {
  content: "Table of Contents";
  font-weight: bold;
  font-size: 1.2em;
  display: block;
  color: navy;
  margin-bottom: 10px;
}


div#TOC li {     /* table of content  */
    list-style:upper-roman;
    background-image:none;
    background-repeat:none;
    background-position:0;
}

h1.title {    /* level 1 header of title  */
  font-size: 22px;
  font-weight: bold;
  color: DarkRed;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}

h4.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 15px;
  font-weight: bold;
  font-family: system-ui;
  color: navy;
  text-align: center;
}

h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Gill Sans", sans-serif;
  color: DarkBlue;
  text-align: center;
}

h1 { /* Header 1 - and the author and data headers use this too  */
    font-size: 20px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}

h2 { /* Header 2 - and the author and data headers use this too  */
    font-size: 18px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 16px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 14px;
  font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

/* Add dots after numbered headers */
.header-section-number::after {
  content: ".";

body { background-color:white; }

.highlightme { background-color:yellow; }

p { background-color:white; }

}
```

```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("pander")) {
   install.packages("pander")
   library(pander)
}
if (!require("ggplot2")) {
  install.packages("ggplot2")
  library(ggplot2)
}
if (!require("tidyverse")) {
  install.packages("tidyverse")
  library(tidyverse)
}

if (!require("plotly")) {
  install.packages("plotly")
  library(plotly)
}
####
knitr::opts_chunk$set(echo = TRUE,       # include code chunk in the output file
                      warning = FALSE,   # sometimes, you code may produce warning messages,
                                         # you can choose to include the warning messages in
                                         # the output file. 
                      results = TRUE,    # you can also decide whether to include the output
                                         # in the output file.
                      message = FALSE,
                      comment = NA
                      )  
```
 
 \
 
## **Assignment Objectives** 

* Master the fundamental concepts of point estimation and performance metrics

* Understand the theoretical foundation of the method of moments estimator (MME)

* Implement MME in R, incorporating numerical approximation methods

\

**Use of AI Tools**

**Policy on AI Tool Use**: Students must adhere to the AI tool policy specified in the course syllabus. The direct copying of AI-generated content is strictly prohibited. All submitted work must reflect your own understanding; where external tools are consulted, content must be thoroughly rephrased and synthesized in your own words.

**Code Inclusion Requirement**: Any code included in your essay must be properly commented to explain the purpose and/or expected output of key code lines. Submitting AI-generated code without meaningful, student-added comments will not be accepted.

\

**Log-logistic Distribution**

The log-logistic distribution (also known as the Fisk distribution) is a continuous probability distribution that is particularly useful in contexts where data exhibit non-negative, skewed behavior and where the hazard rate is unimodal (increases to a peak and then decreases). It has been widely used in the areas such as survival analysis and reliability engineering, environmental science, economics, pharmacology, finance and risk management, etc. 

For given shape parameter $\beta$ and scale parameter $\alpha$, the cumulative distribution function

$$
F(x) = \frac{1}{1+(x/\alpha)^{-\beta}}
$$

As an exercise, you can derive the density in the following form

$$
f(x) = \frac{(\beta/\alpha)(x/\alpha)^{\beta-1}}{[1+(x/\alpha)^\beta]^2}, \ \ \text{ for } \ \ x > 0.
$$

After some algebra, we can find the $k$th moment

$$
\mu_k = E[X^k] = \alpha^k B\left(1+\frac{k}{\beta}, 1 - \frac{k}{\beta} \right).
$$

This assignment will focus on finding MME of parameters $\alpha$ and $\beta$ based on a real-world application data set.


\

## **Question 1: Derive the log-logistic density function **

Given the CDF of the two-parameter log-logistic distribution

$$
F(x) = \frac{1}{1+(x/\alpha)^{-\beta}}.
$$
We find the probability density function by differentiating $F(x)$ with respect to $x$


\newcommand{\der}{\frac{d}{dx}}
\begin{eqnarray*}
    \der F(x) &=& \der \left(1 + \alpha^\beta x^{-\beta}\right)^{-1}\\
    &=& -\left(1 + \alpha^\beta x^{-\beta}\right)^{-2} \der \left(1 + \alpha^\beta x^{-\beta}\right)\\
    &=& -\left(1 + \alpha^\beta x^{-\beta}\right)^{-2} \left(-\alpha^{\beta}\beta x^{-\beta-1}\right)\\
    &=& \frac{\alpha^{\beta}\beta x^{-\beta-1}}{\left(1 + \alpha^\beta x^{-\beta}\right)^{-2}}\\
\end{eqnarray*}

Now, we manipulate our derivative into the form listed above

\begin{eqnarray*}
    \frac{\alpha^{\beta}\beta x^{-\beta-1}}{\left(1 + \alpha^\beta x^{-\beta}\right)^{-2}} &=& \frac{\alpha^{\beta}\beta x^{-\beta-1}}{\left[x^{-\beta} \alpha^{\beta}\left(\frac{x^{\beta}}{\alpha^{\beta}} +1\right)\right]^{2}} \\
    &=& \frac{x^{-\beta}\alpha^{\beta}\beta x^{-1}}{\left[x^{-\beta} \alpha^{\beta}\left(\frac{x^{\beta}}{\alpha^{\beta}} +1\right)\right]^{2}}\\
    &=& \frac{x^{-\beta}\alpha^{\beta}\beta x^{-1}}{x^{-2\beta} \alpha^{2\beta}\left(\frac{x^{\beta}}{\alpha^{\beta}} +1\right)^{2}}\\
    &=& \frac{\beta x^{-1}}{x^{-\beta} \alpha^{\beta}\left(\frac{x^{\beta}}{\alpha^{\beta}} +1\right)^{2}} \\
    &=& \frac{\beta x^{-1}}{\left(\frac{\alpha}{x}\right)^{\beta}\left(\frac{x^{\beta}}{\alpha^{\beta}} +1\right)^{2}}\\
    &=& \left(\frac{\alpha}{x}\right)^{-\beta}\frac{\beta x^{-1}}{\left(\frac{x^{\beta}}{\alpha^{\beta}} +1\right)^{2}}\\    
    &=& \left(\frac{x}{\alpha}\right)^{\beta}\frac{\beta x^{-1}}{\left[\left(\frac{x}{\alpha}\right)^{\beta} +1\right]^{2}}\\
    &=& \frac{ \left(\frac{x}{\alpha}\right)^{\beta} \beta x^{-1}}{\left[\left(\frac{x}{\alpha}\right)^{\beta} +1\right]^{2}}\\
    &=& \frac{\left(\frac{\alpha^{-1}}{\alpha^{-1}}\right) \left(\frac{x}{\alpha}\right)^{\beta} \beta x^{-1}}{\left[\left(\frac{x}{\alpha}\right)^{\beta} +1\right]^{2}}\\
    &=& \frac{\left(\frac{x}{\alpha}\right)^{\beta- 1} \left(\frac{\beta}{\alpha}\right) }{\left[\left(\frac{x}{\alpha}\right)^{\beta} +1\right]^{2}}
\end{eqnarray*}


## **Question 2: Distribution of Recovery Time from A Surgery**

Time to recovery (in days) after a specific knee surgery procedure. This follows a typical **log-logistic pattern** in medical survival/recovery analysis:

```
8.23, 12.74, 14.83, 16.61, 18.16, 19.55, 20.80, 21.94, 23.00, 23.98, 24.89, 25.75, 26.56, 
27.34, 28.08, 28.79, 29.48, 30.15, 30.81, 31.45, 32.08, 32.70, 33.31, 33.92, 34.53, 35.13, 
35.73, 36.33, 36.93, 37.53, 38.14, 38.75, 39.37, 40.00, 40.64, 41.29, 41.95, 42.63, 43.33, 
44.05, 44.79, 45.56, 46.36, 47.20, 48.08, 49.02, 50.03, 51.12, 52.32, 53.65
```
Based on the above data to perform the following analysis.

a) Using method of moment estimation to estimate $\alpha$ and $\beta$, denoted by $\skew{3}\hat{\alpha}$ and $\skew{3}\hat{\beta}$, respectively. 

We have two parameters we wish to find, so we will use a system of equations.

$$
\mu_1 = E[X^1] = \alpha B\left(1+\beta^{-1}, 1 - \beta^{-1} \right)
$$
$$
\mu_2 = E[X^2] = \alpha^2 B\left(1+2\beta^{-1}, 1 - 2\beta^{-1} \right)
$$

squaring $\mu_1$ we get

$$
\mu_{1}^2 = E^2 \left[X^1\right] = \alpha^2 B^2\left(1+\beta^{-1}, 1 - \beta^{-1} \right)
$$
$$
\frac{\mu_{1}^2}{\mu_2} = \frac{B^2\left(1+\beta^{-1}, 1 - \beta^{-1} \right)}{B\left(1+2\beta^{-1}, 1 - 2\beta^{-1} \right)}
$$
and we can find $\alpha$ from

$$
\alpha = \frac{\mu_1}{B\left(1+\beta^{-1}, 1 - \beta^{-1} \right)}
$$

```{r}
X <- c(8.23, 12.74, 14.83, 16.61, 18.16, 19.55, 20.80, 21.94, 23.00, 23.98, 24.89, 25.75, 26.56, 
27.34, 28.08, 28.79, 29.48, 30.15, 30.81, 31.45, 32.08, 32.70, 33.31, 33.92, 34.53, 35.13, 
35.73, 36.33, 36.93, 37.53, 38.14, 38.75, 39.37, 40.00, 40.64, 41.29, 41.95, 42.63, 43.33, 
44.05, 44.79, 45.56, 46.36, 47.20, 48.08, 49.02, 50.03, 51.12, 52.32, 53.65)


moment_estimator <- function(X, k){
  n <- length(X)
  (1/n) * sum(X^k)
}

m_1 <- moment_estimator(X, 1)
m_2 <- moment_estimator(X, 2)

eq_1 <- function(t){
  num <- beta(1 + t^-1, 1-t^-1)^2
  den <- beta(1+2*t^-1, 1-2*t^-1)
  (m_1^2 / m_2) - (num / den)
}



find_bounds <- function(func){
  #wikipedia says that the beta function B(z_1,z_2) is
  #defined for z_1, z_2 > 0. This means 1- 2*beta^-1 > 0 => Meaning beta > 2
  max_iter <- 10
  first_bound <- NaN
  f_first_bound <- NaN
  for( i in seq(from=2.1, by=1, length.out=max_iter)){
    f_i <- func(i)
    if(!is.nan(f_i)){
      f_first_bound <- f_i
      first_bound <- i
      break 
    }
  }

  if (is.nan(first_bound)){
    #lower_bound is still Nan. we went through the whole
    #loop without finding a valid value
    stop("could not find first bound")
  }
  upper_bound <- NaN
  
  for(i in (seq(max_iter) + first_bound)){
    f_i <- func(i)
    if (sign(f_i) != sign(f_first_bound)){
      return(c(first_bound, i)) 
    }
  }
  stop("could not find second bound")
}

eq_1_bounds <- find_bounds(eq_1)
#TODO: what to do about the bounds?
#beta_hat <- uniroot(eq_1, c(2.1, 50))$root
beta_hat <- uniroot(eq_1, eq_1_bounds)$root

eq_2 <- function(t){
  den <- beta(1+t^-1, 1-t^-1)
  m_1 / den
}

alpha_hat <- eq_2(beta_hat)

horizontal_points <- seq(min(X) - 1, max(X) + 1, length.out=length(X))

loglogistic_pdf <- function(X, alpha, beta){
  (beta / alpha) * (X / alpha)^(beta- 1) / (1 + (X / alpha)^beta)^2 
}


data <- data.frame(x=horizontal_points, y=loglogistic_pdf(horizontal_points, alpha_hat, beta_hat))
plot <- ggplot(data=data) + 
  geom_histogram(aes(x=X, y=after_stat(density), color="sample density"), alpha=0, bins=14) + 
  geom_density(aes(x=X, color="gaussian kernel estimate")) + 
  geom_line(aes(x=x, y=y, color="estimated density")) + 
  guides(alpha="none") + 
  labs(title="Comparison of Estimated Density", x="X", y="Density") + 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
  
ggplotly(plot)

```

It appears that our estimated log logistic density assigns more density to the center of our graph that our gaussian kernel estimate. Another major difference appears to be the skewedness of the estimates. The MME log logistic estimate is skewed right, while the kernel estimate is skewed left.


b) Since the moment estimates $\skew{3}\hat{\alpha}$ and $\skew{3}\hat{\beta}$ are random, construct bootstrap sampling distributions for each. To visualize these distributions, plot separate bootstrap histograms for $\skew{3}\hat{\alpha}$ and $\skew{3}\hat{\beta}$.  Then, overlay a smooth density curve on each histogram using Gaussian kernel density estimation. Finally, describe the patterns of these density curves.

```{r}

MGF <- function(k){
  function(a, b){
    a^k * beta(1 + k*b^-1, 1 - k*b^-1)
  }
}

#get functions for the explicit definition of the first and second moments
M_1 <- MGF(1)
M_2 <- MGF(2)


#find the parameters of log-logistic distribution given sample moments
solve_loglog_params <- function(m_1, m_2){
  eq_1 <- function(t){
    m_1^2 / m_2 - M_1(1, t)^2 / M_2(1, t)
  }
  bounds <- find_bounds(eq_1)
  b <- uniroot(eq_1, interval=bounds)$root
  
  eq_2 <- function(t){
    m_1 / beta(1 + t^-1, 1 - t^-1)
  }
  
  a <- eq_2(b)
  
  c(a, b)
}



moment_estimator <- function(X){
  #compute the sample moments
  n <- length(X)
  m_1 <- (1/n) * sum(X)
  m_2 <- (1/n) * sum(X^2)
  #return alpha and beta for these values of m_1 and m_2
  solve_loglog_params(m_1, m_2)
}


#code from last week's homework
boostrap_sample_iteration <- function(X,  estimator){
  sample_i <- sample(X, size=length(X), replace=TRUE)
  estimator(sample_i)
}

#create replicant_n replicant samples
replicant_n <- 5000

#simplify=TRUE param gives us vector as output
bootstrap_estimator_dist <- replicate(n=replicant_n, expr=boostrap_sample_iteration(X, moment_estimator), simplify=TRUE)
#transpose so we get columns of m_1 and m_2, replicant_n rows
bootstrap_estimator_dist <- data.frame(t(bootstrap_estimator_dist))

colnames(bootstrap_estimator_dist) <- c("a", "b")


#plot of alpha
plot <- ggplot(data=bootstrap_estimator_dist) + 
  geom_histogram(aes(x=a, after_stat(density), color="boostrap density"), alpha=0, bins=14) + 
  geom_density(aes(x=a, color="gaussian kernel estimate")) + 
  guides(alpha="none") + 
  #ggplotly will note render tex code
  labs(title="Bootstrap distribution of alpha hat", x="X", y="Density") + 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))

ggplotly(plot)
```
The bootstrap distribution of $\skew{3}\hat{\alpha}$ appears roughly normal.


```{r}
#plot of beta

plot <- ggplot(data=bootstrap_estimator_dist) + 
  geom_histogram(aes(x=b, after_stat(density), color="bootstrap density"),alpha=0, bins=14) + 
  geom_density(aes(x=b, color="gaussian kernel estimate")) + 
  labs(title="Bootstrap distribution of beta hat", x="X", y="Density") + 
  guides(alpha="none") + 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))

ggplotly(plot)

```

The bootstrap distribution of $\skew{3}\hat{\beta}$ appears skewed right.







